# US building permit

```{r setup}
library(tidyverse)
```

## Data import

Data downloaded from https://www.recenter.tamu.edu/data/building-permits/

```{r}
permits_raw <- read_csv("dataPermit_full.csv", na = "null")      # change any "Null" to "NA"
permits_raw                                                      # print data, to look at
```

* area = metropolitan standard area
* date = month / year (character vector, or strings)
* f1 = 1 family house; f24 = 2-4 families; f5 = 5+ family house
* units = number of buildings; change in units; value = average value of building; valchange = change in value

```{r}
?separate

permits <- permits_raw %>%
  separate(date, c("month", "year"), "/", convert = TRUE) %>%     # `convert` to get m&y integers
  filter(year > 2007) %>%         # last 10yrs. (`separate()` enables this filter, with a `year`)
  select(-ends_with("change"))    # remove others' calculated fields, in case they are wrong
permits
```

## Basic EDA

```{r}
# "...just to get a sense of (like) what's going on in this dataset.
# And a really useful tool for that is just a `count()`"...
permits %>% count(year)              # `n` is ~4500 per year
permits %>% count(area)              # `n` is usually 119 per area...
permits %>% count(area) %>% count(n) # dbl-count: how many had that count of area. [useful trick]

# "So it's just useful, to give me a little idea of what's going on"
```

For this analysis I'll focus on single family homes

```{r}
permits <- permits %>%
  mutate(date = year + (month - 1) / 12)
# v rough, for the plot below. Puts January's datapoint at the '0' position, etc

ggplot(permits, aes(date, f1units)) +
  geom_point()                          # "It's always a good idea just to start out with points"
# ...but not very useful here, as most of the data is 'hidden' in a big black blob at the bottom

ggplot(permits, aes(date, f1units)) +
  geom_line()
# ...also not very useful, as only get one line. Need to tell it how it's grouped...

ggplot(permits, aes(date, f1units)) +
  geom_line(aes(group = area))
# ...helps a little bit (can see some of the big cities), but mostly big black blob again
```

## Focus

Let's (arbitrarily) focus on big cities. Might bias our results, because maybe big cities are different from small cities in some ways? ["But even taking that into account, I think it's a good place to start, with the places you've got the most data" (and later extend the analysis to look at all of the cities)]

```{r}
f1units <- permits %>%
  group_by(area) %>%                    # "let's start by figuring out which are the big cities"
  summarise(mean = mean(f1units)) %>%   # avg number of single family dwellings built each month
  arrange(desc(mean))                   # (381 rows)

f1units %>% filter(mean > 100)          # look at cities with (arbitrarily) >100 permits/mth

# "And then now we need to go back and find all the permits for those (96) cities. And I'm going to do that with a semi-join""
```




