# US building permit

```{r setup}
library(tidyverse)
```


## Data import

Data downloaded from https://www.recenter.tamu.edu/data/building-permits/

```{r}
permits_raw <- read_csv("dataPermit_full.csv", na = "null")      # change any "null" to "NA"
permits_raw                                                      # print data, to look at
```

* area = metropolitan standard area (which is how census divides up the country in big cities)
* date = month / year (currently stored as character vector, or strings)
* f1 = 1 family house; f24 = 2-4 families; f5 = 5+ family house
* units = number of buildings; change in units; value = avg value of building; valchange = change in value

```{r}
# "The data goes back quite a long time. So, I'm just going to start off by looking at the last 10 years of data" ... "If I want to filter by year, I really need a `year` variable first"

?separate

permits <- permits_raw %>%
  separate(date, c("month", "year"), "/", convert = TRUE) %>%   # `convert` gets m&y as integers
  filter(year > 2007) %>%       # last 10yrs. (`separate()` enables this filter, with a `year`)
  select(-ends_with("change"))  # to remove others' calculated fields, in case they are wrong
permits                         # (44707 rows)
```


## Basic EDA [@5'00]

```{r}
# "...just to get a sense of (like) what's going on in this dataset. And a really useful tool for that is just a `count()`"...

permits %>% count(year)              # `n` is ~4500 per year
permits %>% count(area)              # `n` is usually 119 per area...
permits %>% count(area) %>% count(n) # dbl-count: qty having that count of area. [useful trick]

# "So it's just useful, to give me a little idea of what's going on"
```

For this analysis I'll focus on single family homes (`f1units`)

```{r}
permits <- permits %>%
  mutate(date = year + (month - 1) / 12)
# v rough, for the plot below. Puts January's datapoint at the '0' position, etc

ggplot(permits, aes(date, f1units)) +
  geom_point()                          # "It's always a good idea just to start out with points"
# ...but not very useful here, as most of the data is 'hidden' in a big black blob at the bottom

ggplot(permits, aes(date, f1units)) +
  geom_line()
# ...also not very useful, as only get one line. Need to tell it how it's grouped...

ggplot(permits, aes(date, f1units)) +
  geom_line(aes(group = area))
# ...helps a little bit (can see a few of the bigger cities), but mostly big black blob again
```


## Focus [@7'45]

Let's (arbitrarily) focus on big cities. Might bias our results, because maybe big cities are different from small cities in some ways? ["But even taking that into account, I think it's a good place to start, with the places you've got the most data" (and later extend the analysis to look at all of the cities)]

```{r}
f1units <- permits %>%
  group_by(area) %>%                    # "let's start by figuring out which are the big cities"
  summarise(mean = mean(f1units)) %>%   # avg number of single family dwellings built each month
  arrange(desc(mean))                   # (381 rows)

f1units %>% filter(mean > 100)          # look at cities with (arbitrarily) >100 permits/mth

# "And then now we need to go back and find all the permits for those (96) cities. And I'm going to do that with a `semi_join`..."

permits_big <- permits %>%
  semi_join(f1units %>% filter(mean > 100)) # 'give me all the permits that match these cities'

permits_big                             # (11424 rows)
```

[@10'00]

"For now, kind of focus on this process : I try out something, I look at the result, and then see how that's going(?)"
"And let's go up and redo this plot. So now, instead of all of the permits data, I'm just going to look at the permits data for the bigger cities"

```{r}
ggplot(permits_big, aes(date, f1units)) +
  geom_line(aes(group = area), alpha = 0.1) # 'one quick trick is to make the lines transparent'

ggplot(permits_big, aes(date, f1units)) +
  geom_line(aes(group = area), alpha = 0.1) +
  scale_y_log10()                           # "maybe I could try doing a log10 transformation"

"Looks like ... maybe there's actually (like) some kind of common pattern going on within a year. And then maybe there's some sort of common long-term trends. One easy way to make those long-term trends a little easier to see is to add on a smoothed line..."

ggplot(permits_big, aes(date, f1units)) +
  geom_line(aes(group = area), alpha = 0.1) +
  scale_y_log10() +
  geom_smooth(se = FALSE)
```

"But it's still a little hard to see, because we've got these 2 problems : That we've got the strong pattern within the year, and the cities are still kind of spread out"


## Model [@12'00]

"Can we do better? And to do that, we're going to use a model. Because models are really powerful tools for partitioning the signal. So we're going to try and partition this pattern - this signal - into this monthly component and then what's left, maybe this smoother long-term trend"

"And it's going to be easiest to do that if I start by focusing on just one city..."

```{r}
library(stringr)                 # (no mention in video, but needed for `str_detect`, just below)

houston <- permits %>% filter(str_detect(area, "Houston"))
houston %>% count(area)                           # (119 rows)

ggplot(houston, aes(date, f1units)) +             # (as for previous chunk, but just Houston)
  geom_line(aes(group = area))                    # (`group` not needed here)

ggplot(houston, aes(month, f1units)) +            # emphasise 'pretty strong seasonal trend'...
  geom_line(aes(group = year)) +                  # ...by separate line for each year
  scale_y_log10()                                 # (this 'log' scaling isn't helpful this time)
```

[@13'30]

"There's some evidence of a strong seasonal pattern here"

Questions :

* Is this pattern the same everywhere?
* What drives it? Is it the weather?
* Houston in July is less pleasant than Houston in December

[@14'30]

"Let's try and extract that pattern out. We're going to do this by using a linear model..."

```{r}
library(modelr)

# "I'm going to try and predict the number of units based on just the month. And I'm going to treat that month, rather than a continuous value [sic], but as a categorical variable"

houston_mod <- lm(log(f1units) ~ factor(month), data = houston) # create a model of Houston

# "...I don't believe this model is a good model (right?). I know there's other stuff going on here. I'm just using it as a way to (kind of) partition the signal into this monthly effect" [@15'04]

houston %>%
  add_predictions(houston_mod)  %>%   # "to add a column [`pred`] of predictions to my dataframe"
  ggplot(aes(date, pred)) +
  geom_line()

# "...And you can see that what this model's captured is just this repeating seasonal pattern" ... "What might be more useful is to look at the residuals from the model"...

# [@16'00]

```


* What's driving this trend?
* What happened around 2010?

## Extend the model to every city [@17'00]

```{r}

```

"much much easier to see the long-term trend, now that we've removed those short-term"
"So let's just quickly recap what we're doing in this data analysis"

[@19'40]






